---
title: "p8105_hw2_zl2746"
author: "Ze Li"
date: "2023-09-27"
output: github_document
---

```{r}
library(tidyverse)
library(readr)
library(readxl)
```

## Problem 1

### Step 1. 

Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`; 

replace month number with month name; 

create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; 

and remove the `day` variable.

```{r p1step1}
month_df = 
  read_csv("fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names()
  
month_sep_df =
  separate(month_df,col=mon,c("year","month","day"),sep="-") |>
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    day = as.integer(day),
    month = month(month, label = TRUE),
    #month_full = month(month, label = TRUE, abbr = FALSE)
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")
  ) |>
  select(-starts_with("day"))
```

### Step 2. 

Clean the data in snp.csv using a similar process to the above. 

For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r p1step2}
snp_df = 
  read_csv("fivethirtyeight_datasets/snp.csv") |>
  janitor::clean_names()

snp_sep_df =
  separate(snp_df,col=date,c("month","day","year"), sep="/") |>
  mutate(
    month = as.integer(month),
    day = as.integer(day),
    year = as.integer(year),
    month = month(month, label = TRUE),
    year = ifelse(year<16, year+2000, year+1900)
  ) |>
  select(-starts_with("day")) |>
  relocate(year)
```

### Step 3. 

Tidy the unemployment data so that it can be merged with the previous datasets. 

This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; 

and ensuring that key variables take the same values.

```{r p1step3}
unem_df = 
  read_csv("fivethirtyeight_datasets/unemployment.csv") |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month",
    values_to = "unemployment"
  ) |> 
  rename(year = Year) |> 
  select(year, month, unemployment) #|>
```

### Step 4. 

Join the datasets by merging snp into pols, and merging unemployment into the result. 

```{r p1step4}
cdata_p1 = 
  left_join(month_sep_df, snp_sep_df) |>
  left_join(x = _, y = unem_df) 

```

### Step 5.

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

The `pol_month` data has `r nrow(month_sep_df)` observations and `r ncol(month_sep_df)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r range(month_sep_df$year)[1]` to `r range(month_sep_df$year)[2]`. It also tells us whether the sitting president was a democrat or republican. 

The `snp` data has `r nrow(snp_sep_df)` observations and `r ncol(snp_sep_df)` variables, ranging from years `r range(snp_sep_df$year)[1]` to `r range(snp_sep_df$year)[2]`. 

The `unemployment` data has `r nrow(unem_df)` observations and `r ncol(unem_df)` variables ranging from years `r range(unem_df$year)[1]` to `r range(unem_df$year)[2]`. 
In Januarys in or after 1975 in which a democrat was president, the **average unemployment rate was `r filter(cdata_p1, month == "Jan", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`**.  

The average unemployment rate over the same time period in which a republican was president was `r filter(cdata_p1, month == "Jan", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.


## Problem 2

```{r problem2}
Mr.Trash_df = 
  read_excel("202309 Trash Wheel Collection Data.xlsx",sheet = 1, range = "A2:N549") |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (weight_tons*500)/30
  ) |>
  add_column(type = "Mr. Trash")

Prof.Trash_df = 
  read_excel("202309 Trash Wheel Collection Data.xlsx",sheet = 2) |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (weight_tons*500)/30,
    year = as.character(year)
  ) |>
  drop_na() |>
  add_column(type = "Professor Trash") 

G.Trash_df = 
  read_excel("202309 Trash Wheel Collection Data.xlsx",sheet = 4) |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (weight_tons*500)/30,
    year = as.character(year)
  ) |>
  drop_na() |>
  add_column(type = "Gwynnda Trash")

cdata_p2 = 
  bind_rows(Mr.Trash_df, Prof.Trash_df) |>
  bind_rows(x = _, y = G.Trash_df)
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables.

The `Mr.Trash` data has `r nrow(Mr.Trash_df)` observations and `r ncol(Mr.Trash_df)` variables and tells us about the  information on the dumpter number, date of collection, amount of total litter and litter type. It also tells Mr.Transh  transferred `r sum(Mr.Trash_df$homes_powered)` power into make electricity for Maryland homes.

The `Prof.Trash` data has `r nrow(Prof.Trash_df)` observations and `r ncol(Prof.Trash_df)` variables and tells us about the  information on the dumpter number, date of collection, amount of total litter and litter type. It also tells Prof.Transh  transferred `r sum(Prof.Trash_df$homes_powered)` power into make electricity.

The `G.Trash` data has `r nrow(G.Trash_df)` observations and `r ncol(G.Trash_df)` variables and tells us about the  information on the dumpter number, date of collection, amount of total litter and litter type. It also tells Gwynnda Transh transferred `r sum(G.Trash_df$homes_powered)` power into make electricity.

For available data, what was the total weight of trash collected by Professor Trash Wheel? 

The total weight of trash collected by Professor Trash Wheel is `r sum(Prof.Trash_df$weight_tons)` tons.

What was the total number of cigarette butts collected by Gwynnda in July of 2021?

The total number of cigarette butts collected by Gwynnda in July of 2021 is `r sum(G.Trash_df$cigarette_butts)`.

## Problem 3

### Step 1. 

Import, clean, and tidy the dataset of baseline demographics. 

Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). 

```{r p3step1}
base_df = 
  read_csv("data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  mutate(
    sex = recode(sex, "0" = "female", "1" = "male"),
    apoe4 = recode(apoe4, "0" = "non-apoe4 carrier", "1" = "apoe4 carrier"),
    age_at_onset = ifelse(age_at_onset == '.', NA, age_at_onset)
  ) |>
  filter(current_age < age_at_onset | is.na(age_at_onset)) 
```

Discuss important steps in the import process and relevant features of the dataset. 

There are  `r nrow(base_df)` were recruited, and of __ these develop MCI.

The average baseline age is `r mean(base_df$current_age)`.

What proportion of women in the study are APOE4 carriers?

### Step 2. 

Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; 

```{r p3step2}
amyloid = read_csv("data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |>
  rename(
    id = study_id,
    time_0 = baseline
    ) |>
  pivot_longer(
    time_0:time_8, 
    names_to = "timeperiod",
    values_to = "years"
  ) 
```

Comment on the steps on the import process and the features of the dataset.

```{r}
diffdata1 =
  anti_join(base_df, amyloid)

diffdata2 =
  anti_join(amyloid, base_df)

cdata_p3 = 
  inner_join(base_df, amyloid)
```

