---
title: "p8105_hw2_zl2746"
author: "Ze Li"
date: "2023-09-27"
output: github_document
---

```{r}
library(tidyverse)
library(readr)
library(readxl)
```

## Problem 1

### Step 1. 

Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`; 

replace month number with month name; 

create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; 

and remove the `day` variable.

```{r p1step1}
month_df = 
  read_csv("fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names()
  
month_sep_df =
  separate(month_df,col=mon,c("year","month","day"),sep="-") |>
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    day = as.integer(day),
    month = month(month, label = TRUE),
    #month_full = month(month, label = TRUE, abbr = FALSE)
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")
  ) |>
  select(-starts_with("day"))
```

### Step 2. 

Clean the data in snp.csv using a similar process to the above. 

For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r p1step2}
snp_df = 
  read_csv("fivethirtyeight_datasets/snp.csv") |>
  janitor::clean_names()

snp_sep_df =
  separate(snp_df,col=date,c("month","day","year"), sep="/") |>
  mutate(
    month = as.integer(month),
    day = as.integer(day),
    year = as.integer(year),
    month = month(month, label = TRUE),
    year = ifelse(year<16, year+2000, year+1900)
  ) |>
  select(-starts_with("day")) |>
  relocate(year)
```

### Step 3. 

Tidy the unemployment data so that it can be merged with the previous datasets. 

This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; 

and ensuring that key variables take the same values.

```{r p1step3}
unem_df = 
  read_csv("fivethirtyeight_datasets/unemployment.csv") |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month",
    values_to = "unemployment"
  ) |> 
  rename(year = Year) |> 
  select(year, month, unemployment) #|>
# drop_na?
```

### Step 4. 

Join the datasets by merging snp into pols, and merging unemployment into the result. 

```{r p1step4}
cdata_p1 = 
  left_join(month_sep_df, snp_sep_df) |>
  left_join(x = _, y = unem_df) 

```



## Problem 2

```{r problem2}
Mr.Trash_df = 
  read_excel("202207 Trash Wheel Collection Data.xlsx",sheet = 1, range = "A2:N549") |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (weight_tons*500)/30
  ) |>
  add_column(type = "Mr. Trash")

Prof.Trash_df = 
  read_excel("202207 Trash Wheel Collection Data.xlsx",sheet = 2) |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (weight_tons*500)/30,
    year = as.character(year)
  ) |>
  drop_na() |>
  add_column(type = "Professor Trash") 

G.Trash_df = 
  read_excel("202207 Trash Wheel Collection Data.xlsx",sheet = 4) |>
  janitor::clean_names() |>
  mutate(
    homes_powered = (weight_tons*500)/30,
    year = as.character(year)
  ) |>
  drop_na() |>
  add_column(type = "Gwynnda Trash")

cdata_p2 = 
  bind_rows(Mr.Trash_df, Prof.Trash_df) |>
  bind_rows(x = _, y = G.Trash_df)
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables.

For available data, what was the total weight of trash collected by Professor Trash Wheel? 

What was the total number of cigarette butts collected by Gwynnda in July of 2021?



## Problem 3

### Step 1. 

Import, clean, and tidy the dataset of baseline demographics. 

Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). 

```{r p3step1}
base_df = 
  read_csv("data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  mutate(
    sex = recode(sex, "0" = "female", "1" = "male"),
    apoe4 = recode(apoe4, "0" = "non-apoe4 carrier", "1" = "apoe4 carrier"),
    age_at_onset = ifelse(age_at_onset == '.', NA, age_at_onset)
  ) |>
  filter(current_age < age_at_onset | is.na(age_at_onset)) 
```



### Step 2. 

Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; 

```{r p3step2}
amyloid = read_csv("data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |>
  rename(
    id = study_id,
    time_0 = baseline
    ) |>
  pivot_longer(
    time_0:time_8, 
    names_to = "timeperiod",
    values_to = "years"
  ) 
```

Comment on the steps on the import process and the features of the dataset.

```{r}
diffdata1 =
  anti_join(base_df, amyloid)

diffdata2 =
  anti_join(amyloid, base_df)

cdata_p3 = 
  inner_join(base_df, amyloid)
```

