---
title: "p8105_hw2_zl2746"
author: "Ze Li"
date: "2023-09-27"
output: github_document
---

```{r}
library(tidyverse)
library(readr)
library(readxl)
```

## Problem 1

Step 1. Use `separate()` to break up the variable `mon` into integer variables `year`, `month`, and `day`; 

replace month number with month name; 

create a `president` variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; 

and remove the `day` variable.

```{r step1}
month_df = 
  read_csv("fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names()
  
month_sep_df =
  separate(month_df,col=mon,c("year","month","day"),sep="-") |>
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    day = as.integer(day),
    month = month(month, label = TRUE),
    #month_full = month(month, label = TRUE, abbr = FALSE)
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")
  ) |>
  select(-starts_with("day"))
```

Step 2. Clean the data in snp.csv using a similar process to the above. 

For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r step2}
snp_df = 
  read_csv("fivethirtyeight_datasets/snp.csv") |>
  janitor::clean_names()

snp_sep_df =
  separate(snp_df,col=date,c("month","day","year"), sep="/") |>
  mutate(
    month = as.integer(month),
    day = as.integer(day),
    year = as.integer(year),
    month = month(month, label = TRUE),
    year = ifelse(year<16, year+2000, year+1900)
  ) |>
  select(-starts_with("day")) |>
  relocate(year)
```

Step 3. Tidy the unemployment data so that it can be merged with the previous datasets. 

This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; 

and ensuring that key variables take the same values.

```{r step3}
unem_df = 
  read_csv("fivethirtyeight_datasets/unemployment.csv") |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month",
    values_to = "unemployment"
  ) |> 
  rename(year = Year) |> 
  select(year, month, unemployment)
```

Step 4. Join the datasets by merging snp into pols, and merging unemployment into the result. 

```{r}
data_3 = 
  left_join(month_sep_df, snp_sep_df) |>
  left_join(x = _, y = unem_df)
```

Step 5. Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).




## Problem 2

```{r}
Mr.Trash_df = 
  read_excel("202207 Trash Wheel Collection Data.xlsx",sheet = 1, range = "A2:N549") |>
  janitor::clean_names() #|>
  #mutate(
    #homes_powered = 
  #)

Prof.Trash_df = 
  read_excel("202207 Trash Wheel Collection Data.xlsx",sheet = 2) |>
  janitor::clean_names()

G.Trash_df = 
  read_excel("202207 Trash Wheel Collection Data.xlsx",sheet = 4) |>
  janitor::clean_names()
```
